Trap spaces of Boolean networks are conflict-free siphons of their Petri net encoding

### 2023-03-11 manuscripts/TCS2022/tcs2022.tex
Justification for considering special types of trap spaces

In the field of systems biology, biologists may want to compute more special types of trap spaces beyond minimal trap spaces~\cite{klarner2017pyboolnet}, which also play crucial roles in analysis of Boolean networks.
We shall show that our proposed methods can be easily adjusted to compute popular types of trap spaces.
We illustrate the adjustments via the ASP-based method (see Subsection~\ref{subsec:computation_asp}), but these adjustments are completely applicable for other approaches such as MaxSAT, CP, and ILP\@.

@@

Second, we consider fixed points in Boolean networks.
Let \(s\) be a fixed point of a Boolean network \(\mathcal{N}\).
We have a subspace \(m\) corresponding to \(s\) as follows: \(\forall v \in V, m(v) = s(v)\), i.e., all nodes are fixed in \(m\).
Clearly, \(s\) is a trap set of \(\mathcal{N}\) regardless of the update scheme.
Hence, \(m\) is a trap space of \(\mathcal{N}\).
In addition, since \(\left\vert S_{\mathcal{N}}[m]\right\vert = 1\), \(m\) is also a minimal trap space.
To compute all fixed points of \(\mathcal{N}\), we can add more constraints to the encoded ASP characterizing all conflict-free siphons (equivalently trap spaces).
For every \(v \in V\), we add to the encoded ASP the rule
\[
  \verb|p-v; n-v.|
\]
that ensures that for every conflict-free siphon \(S\), it contains either \(\verb|p-v|\) or \(\verb|n-v|\) for every \(v \in V\).
Equivalently, the trap space corresponding to \(S\) is always a fixed point.
Now, the set of answer sets of the encoded ASP is equivalent to the set of fixed points of \(\mathcal{N}\).
In particular, when solving the encoded ASP using an ASP solver, we do not need to use the built-in option for computing set-inclusion maximal answer sets.
Note that we can also build another ASP characterizing all fixed points of \(\mathcal{N}\) based on the equivalence between a fixed point of \(\mathcal{N}\) and a deadlock of its Petri net encoding~\cite{chatain2014characterization}.
This approach may give a more compact ASP\@.

Third, we consider the trap spaces intersecting a given subspace \(m^{*}\) of a Boolean network.
A trap space \(m\) intersects \(m^{*}\) if and only if \(S_{\mathcal{N}}[m] \cap S_{\mathcal{N}}[m^{*}] \neq \emptyset\).
It follows that for every \(v\), if \(m^{*}(v) = 0\) then \(m(v) = 0\) or \(m(v) = \star\), if \(m^{*}(v) = 1\) then \(m(v) = 1\) or \(m(v) = \star\).
For the former case, we add to \(\mathcal{L}\) the ASP rule
\[
  \verb|:- n-v.|
\]
that ensures that \(m(v)\) cannot be 1.
For the latter case, we add to \(\mathcal{L}\) the ASP rule
\[
  \verb|:- p-v.|
\]
that ensures that \(m(v)\) cannot be 0.
Now \(\mathcal{L}\) characterizes all trap spaces that intersect \(m^{*}\).

Finally, we consider the trap spaces that are inside a given subspace \(m^{*}\) of a Boolean network.
We first adjust \(\mathcal{L}\) to characterize all such trap spaces.
A trap space \(m\) is inside \(m^{*}\) if and only if \(m(v) = m^{*}(v)\) for every \(v \in D_{m^{*}}\).
If \(m^{*}(v) = 0\), we add to \(\mathcal{L}\) the ASP rule
\[
  \verb|p-v.|
\]
that ensures that \(m(v) = 0\).
If \(m^{*}(v) = 1\), we add to \(\mathcal{L}\) the ASP rule
\[
  \verb|n-v.|
\]
that ensures that \(m(v) = 1\).
It is noted that if we want to compute maximal trap spaces inside \(m^{*}\), we need to exclude the conflict-free siphon corresponding \(m^{*}\) from the solution space.
Specifically, we need to add to \(\mathcal{L}\) the ASP rule
\[
  \verb|p-v_i1| ; \verb|n-v_i1| ; \dots ; \verb|p-v_ik| ; \verb|n-v_ik|.
\]
where \(\{v_{i_1}, \dots, v_{i_k}\}\) is the set of free nodes of \(m^{*}\).
This rule ensures that \(m \neq m^{*}\).
In the case that \(m^{*} = \varepsilon\), we have all maximal trap spaces of the original Boolean network.

>>>>

In the field of systems biology, biologists may want to compute more special types of trap spaces beyond minimal trap spaces~\cite{klarner2017pyboolnet}, which also play crucial roles in analysis and control of Boolean networks~\cite{cifuentes2020control,Rozum2021}.
We shall show that our proposed methods can be easily adjusted to compute popular types of trap spaces.
We illustrate the adjustments via the ASP-based method (see Subsection~\ref{subsec:computation_asp}) because ASP is declarative by nature, but these adjustments are completely applicable for other approaches such as MaxSAT, CP, and ILP\@.

@@

Second, we consider fixed points in Boolean networks.
To date, the analysis of the fixed points of a Boolean network remains
a very useful tool in understanding the behavior of complex biological models not only due to the fact that in some cases the full computation of complex attractors remains intractable, but also because for many biological systems, the expected long-term behavior is not cyclic~\cite{corral2021interplay}.
Furthermore, the fixed point computation is also the crucial starting point for several state-of-the-art methods for computing complex attractors of Boolean networks~\cite{DBLP:conf/bcb/TrinhHB22}.
Let \(s\) be a fixed point of a Boolean network \(\mathcal{N}\).
We have a subspace \(m\) corresponding to \(s\) as follows: \(\forall v \in V, m(v) = s(v)\), i.e., all nodes are fixed in \(m\).
Clearly, \(s\) is a trap set of \(\mathcal{N}\) regardless of the update scheme.
Hence, \(m\) is a trap space of \(\mathcal{N}\).
In addition, since \(\left\vert S_{\mathcal{N}}[m]\right\vert = 1\), \(m\) is also a minimal trap space.
To compute all fixed points of \(\mathcal{N}\), we can add more constraints to the encoded ASP characterizing all conflict-free siphons (equivalently trap spaces).
For every \(v \in V\), we add to the encoded ASP the rule
\[
  \verb|p-v; n-v.|
\]
that ensures that for every conflict-free siphon \(S\), it contains either \(\verb|p-v|\) or \(\verb|n-v|\) for every \(v \in V\).
Equivalently, the trap space corresponding to \(S\) is always a fixed point.
Now, the set of answer sets of the encoded ASP is equivalent to the set of fixed points of \(\mathcal{N}\).
In particular, when solving the encoded ASP using an ASP solver, we do not need to use the built-in option for computing set-inclusion maximal answer sets.
Note that we can also build another ASP characterizing all fixed points of \(\mathcal{N}\) based on the equivalence between a fixed point of \(\mathcal{N}\) and a deadlock of its Petri net encoding~\cite{chatain2014characterization}.
This approach may give a more compact ASP\@.

Third, we consider the trap spaces intersecting a given subspace \(m^{*}\) of a Boolean network.
Such trap spaces are used in the trap space-based control method~\cite{cifuentes2020control}.
A trap space \(m\) intersects \(m^{*}\) if and only if \(S_{\mathcal{N}}[m] \cap S_{\mathcal{N}}[m^{*}] \neq \emptyset\).
It follows that for every \(v\), if \(m^{*}(v) = 0\) then \(m(v) = 0\) or \(m(v) = \star\), if \(m^{*}(v) = 1\) then \(m(v) = 1\) or \(m(v) = \star\).
For the former case, we add to \(\mathcal{L}\) the ASP rule
\[
  \verb|:- n-v.|
\]
that ensures that \(m(v)\) cannot be 1.
For the latter case, we add to \(\mathcal{L}\) the ASP rule
\[
  \verb|:- p-v.|
\]
that ensures that \(m(v)\) cannot be 0.
Now \(\mathcal{L}\) characterizes all trap spaces that intersect \(m^{*}\).

Finally, we consider the trap spaces that are inside a given subspace \(m^{*}\) of a Boolean network.
Such trap spaces are used in the iterative procedure of building the succession diagram of a Boolean network~\cite{Rozum2021}, which is hierarchical.
We first adjust \(\mathcal{L}\) to characterize all such trap spaces.
A trap space \(m\) is inside \(m^{*}\) if and only if \(m(v) = m^{*}(v)\) for every \(v \in D_{m^{*}}\).
If \(m^{*}(v) = 0\), we add to \(\mathcal{L}\) the ASP rule
\[
  \verb|p-v.|
\]
that ensures that \(m(v) = 0\).
If \(m^{*}(v) = 1\), we add to \(\mathcal{L}\) the ASP rule
\[
  \verb|n-v.|
\]
that ensures that \(m(v) = 1\).
It is noted that if we want to compute maximal trap spaces inside \(m^{*}\), we need to exclude the conflict-free siphon corresponding \(m^{*}\) from the solution space.
Specifically, we need to add to \(\mathcal{L}\) the ASP rule
\[
  \verb|p-v_i1| ; \verb|n-v_i1| ; \dots ; \verb|p-v_ik| ; \verb|n-v_ik|.
\]
where \(\{v_{i_1}, \dots, v_{i_k}\}\) is the set of free nodes of \(m^{*}\).
This rule ensures that \(m \neq m^{*}\).
In the case that \(m^{*} = \varepsilon\), we have all maximal trap spaces of the original Boolean network.
